Higher Education Synthesized

### 2025-02-17 artifacts/report/msds_data_mining_project_report.tex
update report while working on slides, add slides, and adjust gitignore for refactoring directory structure

\citestyle{acmauthoryear}

@@

\renewcommand{\shortauthors}{Leakakos}

@@

\section{Introduction}
    American higher education saw growing enrollments throughout the latter half of the 1900s and moving into the 200s. This resulted in increases in the number of higher education institutions (HEIs) in the country, including public and private colleges and universities. These HEIs span from major research institutions down to small liberal arts schools. This has also resulted in increases in the number of faculty and staff to support growing enrollments. The general trend in higher education was growth.

@@

    We will take a look at Integrated Postsecondary Education Data System (IPEDS) data from the National Center for Education Statistics (NCES), housed in the U.S. Department of Education's Institute of Education Sciences (IES). NCES gathers, prepares, and makes available data along with providing analysis on education data. IPEDS is a subset of NCES data that focuses on postsecondary or higher education data.

@@

    We will also include a quick review of a few papers related to synthetic data and synthetic tabular data generative modeling. We will utilize synthetic data, but we are not building it from scratch, so we will not dwell on the deep learning implementation details of these papers.

@@

    With a sweep through some research papers out of the way, I did not find any papers directly applicable to what we are looking at in this project. So, the contribution that we provide is to start a discussion around data mining for IPEDS data along with generative modeling for IPEDS data. We build on the ideas in the papers above -- the general approaches to analysis in the earlier papers, and the use of synthetic tabular data for the latter papers.

\section{Proposed Work}
    For me, at this stage, this is the section that I am looking to narrow down a bit (and I am open to suggestions from readers). I have a couple of ideas planned for the work in this project, and I am looking to lean into my background with the architecting of data systems to build out a wider pipeline.

    We are exploring IPEDS data. We went over the value of that data earlier. To snag the data, we will use some Python scripts to download the data from the IPEDS website and store it in a relational database. This database will run locally in Docker, and the configuration files needed to set up the database will exist in this repository, both for the Docker setup as well as for the database work. IPEDS data already comes in a relatively relational form, so there may not be too much work for the initial load into the database.

    From there, we may leverage the database further to provide preprocessing in the database. One easy way to do this is to set up insert triggers to handle the first pass of standardizing the data into a consistent format or doing some early combining and mutating of data before looking at larger exploratory data analysis (EDA). This may work, but we may also want to pull the data out of the database to go through a more standard EDA. But, there is an option here to mimic what we may need in a larger production pipeline with leveraging database tech at this stage.

    Next we go through a full EDA process, this time with a mind towards how we may want to use the data later in the pipeline rather than the earlier EDA based around simple cleanup. This step will happen outside of the database, so we will look at Python scripts here. At first, this is likely easier with the read-eval-print loop (REPL) functionality of a notebook, though we do not want notebooks in our final pipeline until much closer to the end in production. We need the automation power in scripts instead of notebooks.

    After this round of EDA and any further preprocessing we glean from it, we can then think about warehousing. We will shift the data back into the warehouse, likely just matching the new form of the data to what we need for modeling, but we can decide later on the form of the data in the warehouse. For the scope of this project, we will likely end up with a bespoke warehouse built to immediate modeling or analysis needs rather than a broader warehouse set up in something like a full star schema.

    The next step is to explore the outputs of the pipeline. Two directions stand out to me. One is modeling. Especially when working with larger datasets such as those we deal with in data mining, prediction and clustering jump out as obvious candidates, with clustering typically being much more complicated to set up. The second direction is to explore some data mining-specific approaches and algorithms such as similarity measures, apriori, FP-growth, association or rule mining, or others. Honestly, at this stage, any of these could be valid, and I prefer to start working with the data some before deciding on outputs at this stage, so this is left purposely broad.

    One piece I am pulling to the end here is the synthetic data component. I am already wondering if adding in synthetic data is too much. I have been exploring synthetic data in personal and school projects as well as using it at work, so it is fresh on my mind, and I would like to keep testing it out, but it may be overkill for this project. If we do keep synthetic data, then it makes sense to simplify some of the rest of the pipeline to keep more of the scope focused on the synthetic data modeling. This means we probably focus more at the end of the pipeline on evaluating the quality of the synthetic data and demoing how we would pull insights out of the synthetic dataset, but we probably look at just one approach for that last piece instead of going through a heavy refinement of the other outputs.

@@

    I have spent most of the time here focusing on the technical side of the work. I need to dig into the IPEDS data before feeling more comfortable with the questions in terms of meaning that we will explore, but I expect those to center around identifying patterns in clusters of institutions based on institution attributes, enrollment trends, and graduation trends, possibly even dropping either enrollment or graduation data in favor of simplifying the data scope for right now.

\section{Evaluation}
    Until we decide to drop synthetic data, if we do make that decision, the first evaluations we will explore are how similar the synthetic data is to the real data. We will explore the evaluation metrics in the Synthetic Data Vault Project. A couple of candidates that stand out are per-attribute comparisons and machine learning score. Per-attribute comparisons check the same feature in the real and synthetic datasets, calculate metrics such as mean or standard deviation, and determine the quality of the synthetic data based on how close these measures are across the datasets.

    Machine learning score trains classifiers on the real and synthetic datasets and compares the performance of those models. For example, if we use both classifiers to check something like accuracy or F1 score on a held-out test set, we can evaluate how similar the synthetic data is to the real data based on comparing those scores.

@@

    There are any number of other evaluative metrics for synthetic data. We will explore what is available with the project we use for synthetic data and will clarify here as we learn more during experimentation.

@@

    For supervised modeling outputs, we can look at standard metrics such as accuracy, precision, F1 score, and more for classification problems. We can use a metric like mean squared error (MSE) to measure predictive model output. These evaluations give us firm numbers we can use to determine how well our modeling is doing.

    But, I am actually more intrigued by seeing what we can find with unsupervised learning. Admittedly, the outputs for unsupervised learning can be more difficult to work with, especially at this early stage. Clustering is the first unsupervised modeling that comes to mind. I often think of clustering as a much more advanced version of EDA. We are not really sure what patterns exist in the data, so we supercharge standard EDA and see what patterns we uncover when leveraging machine learning. But we do not have a clean evaluation metric with clustering. We will need to interactively explore the clustering outputs, at least at first. At some point, we can determine a set number of clusters and pivot the clustering into a classification approach, but the more interesting piece here for me is the clustering itself.

@@

    While I feel like there are loads of evaluative methods to explore, I am also interested in looking at some of the algorithms from the second data mining course. I have leveraged things like similarity scores in other projects, but I have not experimented with, for instance, FP growth, and it would be interesting to possibly pick a brand new algorithm to plop at the end of the pipeline and figure out what it means to evaluate that algorithm and what it reveals to us.

    So, at this stage, this section is more of a brainstorming of some of the evaluative approaches. This will get refined down in the next iteration after exploring the IPEDS data a bit and determining if we will keep or drop the synthetic data component of the pipeline.

\section{Discussion}
    I tend to drag out projects, taking the opportunity to add increased complexity and explore new items I am not as familiar with. For this project, part of me wants to focus on sticking a bit more in-bounds in terms of my comfort level with data pipelines, only adding in a little extra complexity and seeing if it is feasible to build the entire pipeline out in a shorter amount of time. While spending forever on projects is fun and educational, I do risk getting too comfortable with taking my time. I could benefit from descoping and focusing on getting smaller chunks working earlier and then deciding if it is worth expanding out from there.

    In that vein, I am hoping that the refinement in going from the first to second draft of this report, spending some time to explore the technical and data sides of the project in between, will help reveal scope from this draft that can be cut. It would be ideal if I could use a three or four week timeline and adjust the project to fit within that, but it does mean that what I have laid out above is probably too ambitious for that timelines. But, after writing this all out, the scope does feel pretty big at the moment.

    Currently I have identified some IPEDS data, but I have not written any code or set up the database. My approach will be to get the data through to at least the initial EDA and preprocessing before starting to evaluate the scope after that. I feel comfortable with the earlier and more technical steps in terms of time, so my question marks sit with the longer EDA, the modeling, and, if included, the synthetic data portion.

    The main challenge I see is that the data ends up being much more complicated to work with than expected, so I will need to shift the scope down even further to prioritize working with difficult data. That would not be a bad way for the project to pivot, but it is not what I am planning on at this stage. If the data turns out to be friendly enough to work with, then I think the next big challenge will be figuring out the right analysis or modeling in terms of the meaning of the data. The ideas above are all interesting from a technical standpoint, but the meaning of what we dig into needs to drive why we are using those approaches.

    The general backup plan is to limit each stage of the pipeline to maybe one or two pieces. This may mean we do some simple EDA along with a correlation matrix, we look at a clearer regression or classification problem, or we leverage synthetic data but trim the input dataset to only a few features. All of these are tactics for trimming the scope, likely the biggest risk for the project at the moment.

\section{Conclusion}
    Not much to conclude at the moment. We will fill this in as we start doing the data work of the project. What I expect to see here is that we trimmed down the IPEDS data, scoped the use case for the project to be a bit narrower, decided to explore synthetic data, and picked a simpler overall output. For a surprise finish, we may end up deciding to pull some data out and run it through a separate business intelligence tool like Tableau with the idea that end users of the production pipeline may want to leverage tools such as that for their purposes.

    I expect the next steps to end up being that we incorporate more IPEDS data to answer a broader range of questions. I do not expect radical changes in the pipeline architecture in next iterations of the project, but I do expect additions and expansions.

@@

\section{Technical Details}
    Need some details plz. Leaving this as a placeholder to pull the technical details out of the narrative above but still keep it available for readers.

>>>>

%\citestyle{acmauthoryear}

@@

@@

\section{Introduction}
    American higher education saw growing enrollments throughout the latter half of the 1900s and moving into the 2000s. This resulted in increases in the number of higher education institutions (HEIs) in the country, including public and private colleges and universities. These HEIs span from major research institutions down to small liberal arts schools. This has also resulted in increases in the number of faculty and staff to support growing enrollments. The general trend in higher education was growth.

@@

    We will take a look at Integrated Postsecondary Education Data System (IPEDS) data from the National Center for Education Statistics (NCES), housed in the U.S. Department of Education's Institute of Education Sciences (IES). NCES gathers, prepares, and makes available data along with providing analysis on education data. IPEDS is a subset of NCES data that focuses on postsecondary or higher education data. \cite{ipeds}

@@

    We will also include a quick review of a few papers related to synthetic data and synthetic tabular data generative modeling. We will utilize synthetic data, but we are not building generative models from scratch, so we will not dwell on the deep learning implementation details of these papers.

@@

    With a sweep through some research papers out of the way, I did not find any papers directly applicable to what we are looking at in this project. So, the contribution that we provide is to start a discussion around data mining for IPEDS data along with generative modeling for IPEDS data. We build on the ideas in the papers above -- the general approaches to analysis in the earlier papers, and the use of synthetic tabular data as described in the latter papers.

\section{Proposed Work}

    The general approach for the work in this project is to leverage Python scripts to move information around, a relational database running in Docker for any pieces of the pipeline that would normally use a database, and notebooks as needed for exploratory and visualization purposes.

    To snag the IPEDS data, we will use Python scripts to download the data from the IPEDS website and store it in a relational database, possibly storing the data on disk as a stepping stone. The database will run locally in Docker. IPEDS data already comes in a relatively relational form, so there may not be too much work for the initial load into the database. Much of the data we end up working with in the real world lives in a database, so this recreates that reality here.

    We may leverage the database for preprocessing. We would need to do some manual exploration first in Python, but, after that, we could look at shifting preprocessing into structured query language (SQL), particularly stored as insert triggers in the database. Right now, there is a question about how that would intersect with exploratory data analysis (EDA) that we would end up doing in Python, but we can figure out the specifics as we dig into the data. My thought here is that a production pipeline would be simpler if we could handle more of the preprocessing automatically in the database and not require that someone hook into separate Python scripts, whether automated or needing to be run manually.

    We will need to go through a full EDA process. This step will happen outside of the database, so we will look at Python scripts here. At first, this is likely easier with the read-eval-print loop (REPL) functionality of a notebook, though we do not want notebooks in our final pipeline until much closer to the end in production. We need the automation power in scripts instead of notebooks. We may also learn more from the EDA that we can bake into extra preprocessing, informing what is mentioned in the last paragraph.

    After this round of EDA and any further preprocessing we glean from it, we can then think about warehousing. We will shift the data back into the database, likely just matching the new form of the data to what we need for modeling, but we can decide later on regarding the form of the data in the warehouse. For the scope of this project, we will likely end up with a bespoke warehouse built to immediate modeling or analysis needs rather than a broader warehouse set up in something like a full star schema.

    The next step is to explore the outputs of the pipeline. It occurs to me that there are two competing outputs in this project: synthetic IPEDS data and standard evaluative findings. We will need to find a balance here, so we may limit the amount of regression, classification, or clustering we do to leave space for the synthetic data generative modeling, possibly demoing a smaller example of how we would glean insights out of the synthetic dataset using classification, for example, rather than adding more scope into that outputs after the generative modeling. The final outputs I have a sense will work well with the synthetic IPEDS data are regression for number of enrollments or degree completions, clustering to find groupings of institutions that we should keep an eye on, and classification for putting institutions into those groupings.

    I have been exploring synthetic data in personal and school projects as well as using it at work, so it is fresh on my mind, and I would really like to include it here, but, as said above, it is in tension with what we might see in a standard data mining pipeline. We will need to pull data out of the warehouse, import the synthetic data libraries, run the generative modeling, and then possibly load the synthetic data back into the database. This area has a bit of a question mark as to how much scope it will take up. I feel comfortable with how the generative modeling will work, but we will need to explore performance with IPEDS data in particular.

@@

\section{Evaluation}
    The first evaluations will capture how similar the synthetic data is to the real data. We will explore the evaluation metrics in the Synthetic Data Vault Project, and, on top of those metrics, a couple of extra candidates that stand out are per-attribute comparisons and machine learning score. Per-attribute comparisons check the same feature in the real and synthetic datasets, calculate metrics such as mean or standard deviation, and determine the quality of the synthetic data based on how close these measures are across the datasets. Machine learning score trains classifiers on the real and synthetic datasets and compares the performance of those models. For example, if we use both classifiers to check something like accuracy or F1 score on a held-out test set, we can evaluate how similar the synthetic data is to the real data based on comparing those scores.

@@

    There are any number of other evaluative metrics for synthetic data. We will explore what is available with the Synthetic Data Vault Project that we use and will clarify here as we learn more during experimentation. It is worth calling out here that we could expand the evaluation of the synthetic data itself, taking up more scope from later steps such as regression or classification or such. The risk of not spending enough time evaluating the synthetic data is that we may end up testing out further modeling on an ill-fitting synthetic dataset.

@@

    For supervised modeling outputs, we can look at standard metrics such as accuracy, precision, F1 score, and more for classification problems. We can use a metric like mean squared error (MSE) or adjusted r-squared ($R^2_a$) to measure regression models. These evaluations give us firm numbers we can use to determine how well our modeling is doing. We may stick with supervised learning if we give more space to synthetic data evaluation since my experience is that the degree of difficulty jumps up when we pivot to unsupervised learning

    That said, I am actually more intrigued by seeing what we can find with unsupervised learning. Admittedly, the outputs for unsupervised learning can be more difficult to work with, especially at this early stage. Clustering is the first unsupervised modeling that comes to mind. I often think of clustering as a much more advanced version of EDA. We are not really sure what patterns exist in the data, so we supercharge standard EDA and see what patterns we uncover when leveraging machine learning. But we do not have a clean evaluation metric with clustering. We will need to interactively explore the clustering outputs, at least at first. At some point, we can determine a set number of clusters and pivot the clustering into a classification approach, but the more interesting piece here for me is the clustering itself.

@@

    While I feel like there are loads of evaluative methods to explore, I am also interested in looking at some of the algorithms from the second data mining course. I have leveraged things like similarity scores in other projects, but I have not experimented with, for instance, FP growth, and it would be interesting to possibly pick a brand new algorithm to plop at the end of the pipeline and figure out what it means to evaluate that algorithm and what it reveals to us. But, similar to with unsupervised learning, my desire to explore synthetic data may mean leaving out some of the new algorithms from the second course for now. This would be an area to explore more in the next iteration of the project though.

\section{Discussion}
    My goal for this project is to stick within a 3-4 week limit, if possible. I tend to drag out projects, taking the opportunity to add increased complexity and explore new items I am curious about. But, while spending forever on projects is fun and educational, I do risk getting too comfortable with taking my time. I could benefit from descoping and focusing on getting smaller chunks working earlier and then deciding if it is worth expanding out from there.

    In that vein, I am hoping that this draft is an upper bound for the work in this project. Some areas where I would like to bound the scope a bit would be in the back-and-forth with the database earlier in the pipeline, how much to dig into synthetic data, how wide of a range of questions to explore in the data, and what types of modeling or pattern mining to use at the end of the pipeline. Since I have not started in on the technical work or deeper dives into the data yet, it feels okay to start out more expansively as is done here and look for the next draft to hone in on a more accurate scope.

    I did spend a bit of time looking through different public datasets, a number of those through NCES, ultimately landing on IPEDS, but I still need to spend a bit more time with the chosen IPEDS data. I have a sense the three domains of IPEDS data may get trimmed down to two domains. But gaining clarity on the data itself will take some time.

    The main challenge I see is that the data ends up being much more complicated to work with than expected, and that will take time to reveal itself, particularly in the understanding and preprocessing steps of the pipeline. One goal will be to get to those steps sooner to figure out what it will be like working with this data. The next big challenge I foresee is when we hit modeling, both the generative modeling for the synthetic data and whichever modeling we choose to employ on that synthetic data, though, as said above, the goal is for the latter to remain simpler in this project, leaving space to focus on the synthetic data itself.

    The general backup plan is to simplify the IPEDS dataset down quite a bit. We may even be able to get away with using just one domain of IPEDS data if needed. Since a major goal of this project is to demonstrate synthesizing IPEDS data, we may also only keep data that looks relatively clean to start with. We will also get a sense of which features in the data may be easier to model, again providing a way to simplify the scope to accommodate competing needs.

    To wrap up with an explicit acknowledgment of risks for my projects, the biggets worry is that the scope grows to an unruly size, so there are a number of comments in this report speaking to ways to keep that in check, spanning from stricter time scoping to simplifying the data to sticking with easier-to-implement modeling.

\section{Conclusion}
    There is not much to conclude at the moment. We will fill this in as we start doing the data work of the project. What I expect to see here is that we trimmed down the IPEDS data, scoped the use case for the project to be a bit narrower, spent more time explore synthetic data and its associated generative modeling, and picked a simpler overall output, likely sticking supervised modeling. For a surprise finish, we may end up deciding to pull some data out and run it through a separate business intelligence tool like Tableau with the idea that end users of the production pipeline may want to leverage tools such as that for their purposes, though this would only come into play if we find that the project turns out to be too quick and simple to get through.

    As for future iterations of the project, I expect that the biggest area of grows would be to incorporate more IPEDS data to answer a broader range of questions. I do not expect radical changes in the pipeline architecture in next iterations of the project, but I do expect additions and expansions in order to accommodate particular questions of interest inherent in the new data.

@@

\section{Technical Details}
    Leaving this as a placeholder to pull the technical details out of the narrative above but still keep it available for readers.
